Language Skills in an Ethnically Segmented Labour market: Estonia 1989 -- 2012

### 2018-08-09 paper/ijm.tex
Maybe the very last comit before submission; some missing grammatical articles added; further improved consistency between  response to reviewers and actual text

The bulk of the existing literature analyses the effect of knowledge
of the local majority language for immigrants. It is well known that
language skills are associated with an income premium
between 5 and 35 percent
\citep{Chiswick1995,bleakley+chin2004,shields+price2002,leslie+lindley2001,chiswick+miller2002,Chiswick2010,Chiswick2015, Dustmann2003}. For
instance, \citet{chiswick2008} argues that ``proficient'' speakers
typically earn about 15\% more than ``not proficient'' speakers.
The effect works largely through improving educational attainment
\citep{bleakley+chin2004,rooth+saarela2007native}. Language skills are
also complementary to other types of human capital
\citep{chiswick+miller2007}.
Immigrants who are fluent in the local language also suffer
substantially less from unemployment \citep{shields+price2002, Dustmann2003}.

@@

The paper continues as follows: The next section provides a brief
background of Estonian society and the labour market. In the following
section, we discuss the data and provide some descriptive
statistics. In \cref{sec:method} we describe the econometric
approach, sections \ref{sec:results} and \ref{sec:extensions} present
the results for unemployment probability and wages. The two last
sections, \ref{sec:segregation-analysis} and \ref{sec:discussion}, are
devoted to an additional analysis of labour market segregation, and
concluding discussion.

@@

Such an influx of mainly
Russian-speaking population
led to two \emph{de facto} official languages
by 1970s.
The society was largely divided across language, and the division was
also reflected at workplace \citep{leppik+vihalemm2015JofBaltStud}.
Certain sectors in the economy, such as the armed forces, railways and the
merchant fleet
were dominated by
Russian-speaking workers while more local industries,
such as agriculture and
public administration,
were dominated by ethnic Estonians.
Both communities followed the media in their own respective language, the
universities taught many subjects in both Russian and Estonian,
and service sector was largely bilingual. However, the bilingualism
was mainly one-sided as the bulk of Estonians spoke Russian, but not
the other way around. A particular institution in parts of the Soviet
Union was a segregated school system, where Russian schools existed along the
schools teaching in the local language. Hence, a common education
system, one of the central integration mechanisms in the developed world,
was (and still largely is) missing in Estonia.
The widening use of Russian also
caused increasing concerns about the future of the Estonian
language, in turn leading to an unwillingness to use Russian by
the ethnic Estonians. This further strengthened the language-based segregation.

@@

Our main analysis relies on the Estonian Labour Force Survey (ELFS),
first conducted in 1995 by the Estonian Statistical Office. We limit
ourselves with years 2000--2012 in the main analysis, motivated by a
methodology break: Until 1999 the survey was conducted as an annual
cross-section and later as a quarterly rotating panel. The survey
contains data on approximately 4000 distinct individuals annually,
however, due to its rotating nature, we have roughly 16,000 annual
observations for each year since 2000. For the long-term analysis, we
also use the earlier waves. These waves also include a retrospective
section that contains income up to several years back.

@@

The labour market outcomes we analyse are income, the ``last salary on the main job'' and
labour force status (working, unemployed or inactive).

@@

The central results for unemployment regressions are presented in
\cref{tab:unemployment_estimation_by_sex_and_ethnic_same_sample}\footnote{We
 restrict both models to have the same number of observations as
 requested by a reviewer. The unrestricted results are provided in the Appendix (\Cref{tab:unemployment_estimation_by_sex_and_ethnic}).}.
The
table clearly indicates that Estonian language skills are associated
with an approximately 5 percentage points lower unemployment for both
Russian men and women. We also see that the effect is not 
influenced in any major way by the individual characteristics--for men it is
virtually unchanged while for women the estimate falls by 30\%, from
6.5 to 4.5 percentage points.

@@

		\label{fig:long-run_unemployment_estonian_russian_men}
	}
	\subfigure[Non-native languages for women]{      
		\includegraphics[width=0.45\linewidth]{Figure2b.eps}
		\label{fig:long-run_unemployment_estonian_russian_women}
	}
	\subfigure[English for Men]{
		\includegraphics[width=0.45\linewidth]{Figure2c.eps}
		\label{fig:long-run_unemployment_english_men}
	}
	\subfigure[English for Women]{
		\includegraphics[width=0.45\linewidth]{Figure2d.eps}
		\label{fig:long-run_unemployment_english_women}
	}
	\caption{The effect of language fluency on unemployment, 1992--2012. \\ Control variables are \modelTwo. \agerestrictions }
	\label{fig:long-run_unemployment}
\end{figure}

@@

\begin{figure}[htb]
	\centering
	\subfigure[Non-native languages for men]{
		\includegraphics[width=0.45\linewidth]{Figure3a.eps}
		\label{fig:long-run_wage_estonian_russian_men}
	}
	\subfigure[Non-native languages for women]{
		\includegraphics[width=0.45\linewidth]{Figure3b.eps}
		\label{fig:long-run_wage_estonian_russian_women}
	}
	\subfigure[English for Men]{
		\includegraphics[width=0.45\linewidth]{Figure3c.eps}
		\label{fig:long-run_wage_english_men}
	}
	\subfigure[English for Women]{
		\includegraphics[width=0.45\linewidth]{Figure3d.eps}
		\label{fig:long-run_wage_english_women}
	}
	\caption{The effect of language fluency on wage, 1992--2012. \\ Control variables are \modelTwo. \agerestrictions}
	\label{fig:long-run_wage}
\end{figure}

@@

The previous results can be explained by a combination of two hypotheses: first,
Russian--speaking men, at least those with high school degree as their
highest completed education level, are working
primarily in Russian language environment (ethnic segregation) while this is not true for
women; and second, Russian-speaking men are working in jobs with
little communication requirements (gender segregation).
To get some additional evidence on these hypotheses, we use data
from the 2008 Integration Monitor Survey. We focus on
two questions: ``which is the language of communication at your
workplace?'' and ``which languages do you need at work''. The first
question addresses the ethnic segregation while the second focuses on
the need of the language for communication.

@@

The table indicates that Russian women need Estonian 26 percentage
points more at work than Russian men. The table also suggests that
Russian women are 16 percentage points more likely to be working in
Estonian-speaking environment when controlling for other factors.
Both figures are statistically significant (at 5\% and 10 \% level
respectively), although the standard errors are large. We can also see that
college education is associated with much more Estonian need and usage,
but in regions where large amounts of Russian speakers live, the capital
Tallinn and the North-East, Estonian is less needed. In contrast, no
gender difference is visible for English needs.

@@

We document that all analysed language skills are associated with
unemployment and wage, however in different ways. Our most
intriguing observation is that the largest group of
Russian men, high school graduates, earn virtually no income premium if fluent in Estonian,
the majority and the official language of the country, while ability
to speak the language is related to a substantially
lower unemployment rate.
Our analysis supports \citet{YaoandOurs2015}, and \citet{Lindemann2013} findings that 
the labour market is segmented between ethnic and gender lines.
However, we find that ethnic segregation is less prevalent in the male
labour market.
This suggests for the majority of Russian-speaking men, Estonian
language gives a distinct advantage by facilitating hiring by
Estonian-dominated employers. We also confirm
\citet{leppik+vihalemm2015JofBaltStud} conclusion that
these
jobs are associated with little career mobility. There is also some
anecdotal evidence suggesting that offices are
typically manned by ethnic Estonians while manufacturing units have a
large share of Russian workers.
One can speculate that this outcome is related to a substantial population share
with negative attitudes toward Russians \citep{korts2009JofBaltStud}.
On a more positive tone, we find
that this may not be the case for highly educated workers.

>>>>

The bulk of the existing literature analyses the effect of knowledge
of the local majority language for immigrants. It is well known that
language skills are associated with an income premium
between 5 and 35 percent
\citep{Chiswick1995,bleakley+chin2004,shields+price2002,leslie+lindley2001,chiswick+miller2002,Chiswick2010,Chiswick2015, Dustmann2003}. For
instance, \citet{chiswick2008} argues that ``proficient'' speakers
typically earn about 15\% more than ``not proficient'' speakers.
The effect works largely through improving educational attainment
\citep{bleakley+chin2004,rooth+saarela2007native}. Language skills are
also complementary to other types of human capital
\citep{chiswick+miller2007}.
Immigrants who are fluent in the local ethnic majority language also suffer
substantially less from unemployment \citep{shields+price2002, Dustmann2003}.

@@

The paper continues as follows: The next section provides a brief
background of the Estonian society and the labour market. In the following
section, we discuss the data and provide some descriptive
statistics. In \cref{sec:method} we describe the econometric
approach, sections \ref{sec:results} and \ref{sec:extensions} present
the results for unemployment probability and wages. The two last
sections, \ref{sec:segregation-analysis} and \ref{sec:discussion}, are
devoted to an additional analysis of labour market segregation, and
concluding discussion.

@@

Such an influx of a mainly
Russian-speaking population
led to two \emph{de facto} official languages
by 1970s.
The society was largely divided across language, and the division was
also reflected at workplace \citep{leppik+vihalemm2015JofBaltStud}.
Certain sectors in the economy, such as the armed forces, railways and the
merchant fleet
were dominated by
Russian-speaking workers while more local industries,
such as agriculture and
public administration,
were dominated by ethnic Estonians.
Both communities followed the media in their own respective language, the
universities taught many subjects in both Russian and Estonian,
and service sector was largely bilingual. However, the bilingualism
was mainly one-sided as the bulk of Estonians spoke Russian, but not
the other way around. A particular institution in parts of the Soviet
Union was a segregated school system, where Russian schools existed along the
schools teaching in the local language. Hence, a common education
system, one of the central integration mechanisms in the developed world,
was (and still largely is) missing in Estonia.
The widening use of Russian also
caused increasing concerns about the future of the Estonian
language, in turn leading to an unwillingness to use Russian by
the ethnic Estonians. This further strengthened the language-based segregation.

@@

Our main analysis relies on the Estonian Labour Force Survey (ELFS),
first conducted in 1995 by the Estonian Statistical Office. We limit
ourselves to the years 2000--2012 in the main analysis, motivated by a
methodology break: Until 1999 the survey was conducted as an annual
cross-section and later as a quarterly rotating panel. The survey
contains data on approximately 4000 distinct individuals annually,
however, due to its rotating nature, we have roughly 16,000 annual
observations for each year since 2000. For the long-term analysis, we
also use the earlier waves. These waves also include a retrospective
section that contains income up to several years back.

@@

The labour market outcomes we analyse are income, the ``last salary on the main job'', and
labour force status (working or unemployed).

@@

The central results for unemployment regressions are presented in
\cref{tab:unemployment_estimation_by_sex_and_ethnic_same_sample}\footnote{We
 restrict both models to have the same number of observations as
 requested by a reviewer. The unrestricted results are provided in the Appendix (\Cref{tab:unemployment_estimation_by_sex_and_ethnic}).}.
The
table clearly indicates that Estonian language skills are associated
with an approximately 5 percentage points lower unemployment for both
Russian men and women. We also see that the effect is not 
influenced in any major way by the individual characteristics-- for men it is
virtually unchanged while for women the estimate falls by 30\%, from
6.5 to 4.5 percentage points.

@@

		\label{fig:long-run_unemployment_estonian_russian_men}
	}
	\subfigure[Non-native languages for women]{      
		\includegraphics[width=0.45\linewidth]{Figure2b.eps}
		\label{fig:long-run_unemployment_estonian_russian_women}
	}
	\subfigure[English for Men]{
		\includegraphics[width=0.45\linewidth]{Figure2c.eps}
		\label{fig:long-run_unemployment_english_men}
	}
	\subfigure[English for Women]{
		\includegraphics[width=0.45\linewidth]{Figure2d.eps}
		\label{fig:long-run_unemployment_english_women}
	}
	\caption{The effect of language fluency on unemployment, 1992--2012. \\ Control variables are \modelTwo; \agerestrictions }
	\label{fig:long-run_unemployment}
\end{figure}

@@

\begin{figure}[htb]
	\centering
	\subfigure[Non-native languages for men]{
		\includegraphics[width=0.45\linewidth]{Figure3a.eps}
		\label{fig:long-run_wage_estonian_russian_men}
	}
	\subfigure[Non-native languages for women]{
		\includegraphics[width=0.45\linewidth]{Figure3b.eps}
		\label{fig:long-run_wage_estonian_russian_women}
	}
	\subfigure[English for Men]{
		\includegraphics[width=0.45\linewidth]{Figure3c.eps}
		\label{fig:long-run_wage_english_men}
	}
	\subfigure[English for Women]{
		\includegraphics[width=0.45\linewidth]{Figure3d.eps}
		\label{fig:long-run_wage_english_women}
	}
	\caption{The effect of language fluency on wage, 1992--2012. \\ Control variables are \modelTwo; \agerestrictions}
	\label{fig:long-run_wage}
\end{figure}

@@

The previous results can be explained by a combination of two hypotheses: first,
Russian--speaking men, at least those with high school degree as their
highest completed education level, are working
primarily in a Russian language environment (ethnic segregation) while this is not true for
women; and second, Russian-speaking men are working in jobs with
little communication requirements (gender segregation).
To get some additional evidence on these hypotheses, we use data
from the 2008 Integration Monitor Survey. We focus on
two questions: ``which is the language of communication at your
workplace?'' and ``which languages do you need at work''. The first
question addresses the ethnic segregation while the second focuses on
the need of the language for communication.

@@

The table indicates that Russian women need Estonian 26 percentage
points more at work than Russian men. The table also suggests that
Russian women are 16 percentage points more likely to be working in an
Estonian-speaking environment when controlling for other factors.
Both figures are statistically significant (at the 5\% and 10 \% level
respectively), although the standard errors are large. We can also see that
college education is associated with much more Estonian need and usage,
but in regions where large amounts of Russian speakers live, the capital
Tallinn and the North-East, Estonian is less needed. In contrast, no
gender difference is visible for English needs.

@@

We document that all analysed language skills are associated with
unemployment and wage, however in different ways. Our most
intriguing observation is that the largest group of
Russian men, high school graduates, earn virtually no income premium if fluent in Estonian,
the majority and official language of the country, while ability
to speak the language is related to a substantially
lower unemployment rate.
Our analysis supports \citet{YaoandOurs2015}, and \citet{Lindemann2013} findings that 
the labour market is segmented between ethnic and gender lines.
However, we find that ethnic segregation is less prevalent in the male
labour market.
This suggests for the majority of Russian-speaking men, Estonian
language gives a distinct advantage by facilitating hiring by
Estonian-dominated employers. We also confirm
\citet{leppik+vihalemm2015JofBaltStud} conclusion that
these
jobs are associated with little career mobility. There is also some
anecdotal evidence suggesting that offices are
typically manned by ethnic Estonians while manufacturing units have a
large share of Russian workers.
One can speculate that this outcome is related to a substantial population share
with negative attitudes toward Russians \citep{korts2009JofBaltStud}.
On a more positive tone, we find
that this may not be the case for highly educated workers.
